The (floating) point of homoiconicity

### 2016-06-07 paper/paper.tex
some paper changes, but mostly reorganizing the file system

\usepackage{courier}
\usepackage{listings}
\lstset{
    language=Haskell,
    basicstyle=\ttfamily,
    keywordstyle=\ttfamily,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

@@

\begin{abstract}
%This pearl uses Haskell's type classes to enable a limited form of \emph{homoiconicity}.
%%Specifically, we show how to convert any Haskell98 function that is parametric in its return type into an isomorphic abstract syntax tree.
%Specifically, we show how to convert a certain set of polymorphic functions into isomorphic abstract syntax trees.
%But this is really a pearl about the story of numerical programming in Haskell.

%Numeric programming in Haskell sucks.
%Actually, it sucks in every language because it requires floating point numbers,
%and floating point arithmetic is so difficult to work with.
%The goal of this pearl is to make numeric programming suck just a little bit less.
%This statement is tautologically true---numeric programming requires floating point arithmetic, and working with floats sucks in every language.
%I think this makes numeric programming in Haskell suck just a little bit less.

Floating point numbers are dangerous.
Due to their limited precision,
seemingly correct mathematical formulae can give wildly inaccurate results.
This pearl uses Haskell's type system to mitigate this danger.

The main observation is that many numerical Haskell functions are \emph{homoiconic}.
That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
The construction does not use template Haskell, and so works at both compile and run time.
We use these ASTs to solve two problems with floating point numbers:
\begin{enumerate}
\item
We show how to write functions that automatically stabilize (and optimize!) mathematical formulae.
This lets the programmer write naive floating point code;
the compiler will worry about improving the numeric stability and speed later.
\item
We show how to specify type class laws for floating point numbers that can be easily and automatically tested.
This is a difficult unsolved problem because floating point numbers do not obey many of the traditional algebraic laws like associativity.
\end{enumerate}
We demonstrate these techniques using GHC's builtin numeric hierarchy,
but these techniques generalize to user defined numeric hierarchies as well.

%First, we show how to convert a certain type of commonly used mathematical function into isomorphic abstract syntax trees.
%Specifically, we show that a subset of Haskell functions (which we call \emph{homogeneous}) can easily be converted into an isomorphic abstract syntax tree.

%I hate numerical programming in Haskell,
%but I hate numerical programming in every language.
%It requires floating point numbers,
%which
%
%I hate floating point numbers.

%This pearl provides a technique that makes working with approximate algebraic structures more convenient.
%The prototypical example of an approximate structure is floating point arithmetic, which is famously not a semigroup because it is not associative.
%The technique has two parts.
%First, it lets us construct an FAlgebra (and hence a syntax tree) for a numerical class hierarchy.
%Second, we show how to treat topology as an algebraic structure, and that floating point types have a useful topology.
%This topology can be used to automatically construct properties for approximate algebraic structures that can be easily checked using quickcheck or smallcheck.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

@@

\keywords
floating point, homoiconic, type classes

@@

Imagine you're writing a machine learning library in Haskell.
At some point, you'll need to implement the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic1| will return negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so our function must be rewritten.

Writing numerically stable functions is a notoriously difficult black art.
(See \cite{goldberg1991,higham2002,hamming2012} for some classic introductions.)
Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
First, check our input to see if it would overflow the |exp| function.
If no overflow would occur, then calculate our function as normal;
but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
Performing this transformation gives us:
\begin{lstlisting}
logLogistic2 :: (Floating x, Ord x) => x->x
logLogistic2 x = if x < (-709)
    then x
    else log(1/1+exp(-x))
\end{lstlisting}
%This new function is numerically stable, but we've introduced two new problems:
We've made our function numerically stable, but we've introduced two new problems:

@@

\begin{problem}[specialization]
The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
To properly handle both |Float|s and |Double|s,
we must write two separate functions.
In practice, we may want to support many other numeric types as well.
Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
\end{problem}

These problems are not unique to the log-logistic function.
All functions over the real or complex numbers need to be transformed for floating point stability,
and all such manual transformations suffer from these problems.
%(In fact, the log-logistic transformation described above might be considered an exceptionally simple case.)
%There is a large body of literature on the theory of automating these transformations,
%but relatively little on improving the programmer's experience of dealing with floating point numbers.
%This pearl focuses on the latter.
This pearl presents a method for alleviating these problems.
Specifically, we will see how to write a function
\begin{lstlisting}
stabilize :: (Floating x, Ord x)
    => (x -> x) -> (x -> x)
\end{lstlisting}
that automatically takes an input function like |logLogistic1| and outputs a numerically stable version of that function.
Unfortunately, this introduces a third problem:

@@

\section{Homogeneous Functions are Homoiconic}
Consider the simple numeric class hierarchy shown in Figure \ref{code:group}.
With this class hierarchy, we can write a function that measures the degree to which the |(+)| operation obeys the associative rule.
\begin{lstlisting}
associator :: Group a => a -> a -> a -> a
associator a1 a2 a3
    = ((a1+a2)+a3) - (a1+(a2+a3))
\end{lstlisting}
We call the |associator| function \emph{homogeneous} because:
\begin{enumerate}
\item
The function's return type and all of its parameters are the same type variable.
\item
The constraints contain exactly one type class, and
that type class has exactly one parameter.
\end{enumerate}
Notice that each of the class methods for |Semigroup|, |Monoid|, and |Group| are also homogeneous.
Therefore, we will also say these classes are homogeneous.

@@

\subsection{A type class for FAlgebras}
Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
See for example \cite{swierstra2008}.
The heart of our construction is the |HomFAlgebra| type class, defined as
\begin{lstlisting}
class
    ( Functor (HomSig alg)
    , Foldable (HomSig alg)
    ) => HomFAlgebra (alg::Type->Constraint)
        where
    data HomSig alg a
    runHomSig :: alg a => HomSig alg a -> a
\end{lstlisting}
The |HomFAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
This means that instances of |HomFAlgebra| will be other type classes (instead of types or type constructors).
The |Hom| in |HomFAlgebra| stands for homogeneous.
Most of the names used in this section will be prefixed with |Hom| to differentiate them from related names used in the later heterogeneous constructions.

The easiest way to explain the |HomFAlgebra| class is by example.
We'll start with the |Semigroup| type class defined above.
\begin{lstlisting}
instance HomFAlgebra Semigroup where
    data HomSig Semigroup a = HomSig_plus a a
    runHomSig (HomSig_plus a1 a2) = a1+a2
\end{lstlisting}
The |HomSig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
That is, for every class method, |HomSig| should have a constructor with the same number of parameters.
Since |Semigroup| has a single function that takes two parameters,
the |HomSig Semigroup| data instance has a single constructor that takes two parameters.
The |runHomSig| function evaluates the signature.
For each constructor in |HomSig|, |runHomSig| should apply the corresponding function to the constructor's parameters.

To complete the |HomFAlgebra Semigroup| instance, we must also define the following superclass instances.
\begin{lstlisting}
instance Functor (HomSig Semigroup) where
    fmap f (HomSig_plus a1 a2)
        = HomSig_plus (f a1) (f a2)

instance Foldable (HomSig Semigroup) where
    foldr f b (HomSig_plus a1 a2)
        = f a2 (f a1 b)
\end{lstlisting}
As usual, there is only a single valid |Functor| instance.
There are multiple valid |Foldable| instances depending on the order in which |foldr| considers the constructors' parameters,
but for our purposes the order is not important.

Instances get slightly more complicated when the parameter class has superclasses.
In this case, the |HomSig| data instance must also have a constructor for each superclass that stores the superclass's signature.
This can be seen in the |Monoid| and |Group| instances below.
%The |HomFAlgebra| instances of |Monoid| and |Group| demonstrate this.
\begin{lstlisting}
instance HomFAlgebra Monoid where
    data HomSig Monoid a
        = HomSig_zero
        | HomSig_Semigroup_Monoid
            (HomSig Semigroup a)
    runSig HomSig_zero = zero
    runSig (HomSig_Semigroup_Monoid s)
        = runHomSig s

instance HomFAlgebra Group where
    data HomSig Group a
        = HomSig_negate a
        | HomSig_minus a a
        | HomSig_Monoid
            (HomSig_Monoid_Group a)
    runSig (HomSig_negate a) = a
    runSig (HomSig_minus a1 a2) = a1-a2
    runSig (HomSig_Monoid_Group s)
        = runHomSig s
\end{lstlisting}
Notice that the data instance |HomSig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
The corresponding |Functor| and |Foldable| instances have been omitted for brevity,
but can be found in the paper's git repo.

%The procedure for generating |HomFAlgebra| instances is purely mechanical.
%The ?? library automates the process with the |mkHomFAlgebra| template Haskell function.

@@

We can convert the signature of an FAlgebra into an AST as follows:
\begin{lstlisting}
type HomAST alg a = HomFree (HomSig alg) a

data HomFree f a
    = HomPure a
    | HomFree (f (HomFree f a))
\end{lstlisting}
Intuitively,
the |HomPure| constructor represents a leaf in the AST,
and the |HomFree| constructor represents a branch.
The parameter to |HomFree| is filled by the appropriate |HomSig| data instance.
The constructor used for the |HomSig| instance corresponds to an operation,
and the parameters to the constructor will contain ASTs nested recursively.

For example, consider the expression |(1+2)+3|.
We can create a corresponding AST using |Semigroup|'s |HomFAlgebra| instance.
\begin{lstlisting}
HomFree
  (HomSig_plus
    (HomFree
      (HomSig_plus
        (HomPure 1)
        (HomPure 2)
      )
    )
    (HomPure 3)
  )
\end{lstlisting}
We can also create a corresponding AST for any class that has |Semigroup| as a superclass.
Below is the |Group| AST.
\begin{lstlisting}
HomFree
  (HomSig_Monoid_Group
    (HomSig_Semigroup_Monoid
      (HomSig_plus
        (HomFree
          (HomSig_Monoid_Group
            (HomSig_Semigroup_Monoid
              (HomSig_plus
                (HomPure 1)
                (HomPure 2)
              )
            )
          )
        )
        (HomPure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |HomSig_plus| constructor must be embedded into the |Group|'s |HomSig| with calls to |HomSig_Monoid_Group| and |HomSig_Semigroup_Monoid|.

Constructing these ASTs by hand is a tedious process.
There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |HomSig_plus| constructor.
We can avoid this boilerplate with the following type class.
\begin{lstlisting}
class
    ( HomFAlgebra alg1
    , HomFAlgebra alg2
    ) => HomView alg1 alg2
        where
    embedHomSig
        :: HomSig alg1 a -> HomSig alg2 a
    unsafeExtractHomSig
        :: HomSig alg2 a -> HomSig alg1 a
\end{lstlisting}
Instances of the |HomView| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
For example, |Semigroup| is a superclass of |Group|,
so we should have a |HomView Semigroup Group| instance.
\begin{lstlisting}
instance HomView Semigroup Group
    embedHomSig s
        = HomSig_Monoid_Group (embedHomSig s)
    unsafeExtractHomSig (HomSig_Monoid_Group s)
        = unsafeExtractHomSig s
\end{lstlisting}
Because of this property, we know that the |HomSig alg1 a| data instance can always be embedded into a |HomSig alg2 a| data instance.
But it is only sometimes true that we can extract a |HomSig alg1 a|,
which is why the corresponding function is labeled |unsafe|.

Armed with the |HomView| class, we can create a uniform representation for all class's ASTs.
The expression |(1+2)+3| can now be represented as:
\begin{lstlisting}
HomFree
  (embedHomSig
    (HomSig_plus
      (HomFree
        (embedHomSig
          (HomSig_plus
            (HomPure 1)
            (HomPure 2)
          )
        )
      )
      (HomPure 3)
    )
  )
\end{lstlisting}

%Every FAlgebra can be embedded into itself.
%\begin{lstlisting}
%instance HomView alg alg where
    %embedSig = id
    %unsafeExtractSig = id
%\end{lstlisting}

@@

Using our uniform representation of an AST,
we can make the AST itself an instance of any homogeneous type class.
Again, we'll approach the procedure by example.
We can write a |Semigroup| instance for |HomAST Semigroup a| as:
\begin{lstlisting}
instance
    ( HomView Semigroup alg
    ) => Semigroup (HomFree (HomSig alg) a)
        where
    (+) e1 e2 = HomFree $ embedHomSig $
            HomSig_plus e1 e2
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |HomFree $ embedHomSig $|.
The |Monoid| and |Group| instances follow a similar pattern.
\begin{lstlisting}
instance
    ( HomView Semigroup alg
    , HomView Monoid alg
    ) => Monoid (HomFree (HomSig alg) a)
        where
    zero = HomFree $ embedHomSig $
        HomSig_zero

instance
    ( HomView Semigroup alg
    , HomView Monoid alg
    , HomView Group alg
    ) => Group (HomFree (HomSig alg) a)
        where
    negate e  = HomFree $ embedHomSig $
        HomSig_negate e
    (-) e1 e2 = HomFree $ embedHomSig $
        HomSig_minus e1 e2
\end{lstlisting}
We can easily create ASTs using these instances.
The |Semigroup| AST can be created with the expression
\begin{lstlisting}
(HomPure 1+HomPure 2)+HomPure 3
    :: HomAST Semigroup a
\end{lstlisting}
And the |Group| AST can be created by just changing the type signature.
\begin{lstlisting}
(HomPure 1+HomPure 2)+HomPure 3
    :: HomAST Group a
\end{lstlisting}

\subsection{Showing the AST}
Now that we can create our ASTs, we want to actually do something with them!
The simplest task is to convert them into a |String| for display.
This is an easy procedure of writing |Show| instances for the |HomSig| and |HomFree| data types.

The are no tricks in the |Show| instance for |Semigroup|'s |HomSig|.
\begin{lstlisting}
instance Show a =>
    Show (HomSig Semigroup a)
        where
    show (HomSig_plus a1 a2)
        = show a1++"+"++show a2
\end{lstlisting}
For each operation constructor, we simply show the parameters and the operation connecting them.
For type classes with superclasses, we also need to recursively show the superclasses as can be seen in the |Monoid| and |Group| instances below.
\begin{lstlisting}
instance Show a =>
    Show (HomSig Monoid a)
        where
    show (HomSig_Semigroup_Monoid s)
        = show s
    show (HomSig_zero) = "zero"

instance Show a =>
    Show (HomSig Group a)
        where
    show (HomSig_Monoid_Group s)
        = show s
    show (HomSig_negate a)
        = "negate "++show a
    show (HomSig_minus a1 a2)
        = show a1++"-"++show a2
\end{lstlisting}
The |HomFree| type's |Show| instance is similarly straightforward..
\begin{lstlisting}
instance (Show a, Show (f (HomFree f a)))
    => Show (HomFree f a)
        where
    show (HomPure a) = show a
    show (HomFree f) = "("++show f++")"
\end{lstlisting}
If the syntax tree contains only a single leaf, we just show that leaf.
If the syntax tree contains a branch,
then put parenthesis around the branch so that we don't lose information about the order of operations, and show the functor in the middle.

\subsection{Evaluating the syntax tree}
Another useful task for ASTs is to evaluate them,
generating the value they represent.
This data type is called the free monad because as long as the parameter |f| is a |Functor|,
then |HomFree| has a valid |Monad| instance.
Since all |Monad|s are by necessity also |Functor|s,
|HomFree| has a valid |Functor| instance as well.
\begin{lstlisting}
instance Functor f => Functor (HomFree f) where
    fmap g (HomPure a) = HomPure (g a)
    fmap g (HomFree f)
        = HomFree (fmap (fmap g) f)
\end{lstlisting}
For our purposes, that's all the structure we'll need.

>>>>

\usepackage{courier}
\usepackage{listings}
\lstset{
    language=Haskell,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\ttfamily\footnotesize,
    %basicstyle=\footnotesize,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

@@

\begin{abstract}
Floating point numbers are dangerous.
Due to their limited precision,
seemingly correct mathematical formulae can give wildly inaccurate results.
This pearl shows that Haskell's type system can mitigate this danger.

The main observation is that many numerical Haskell functions are \emph{homoiconic}.
That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
The construction does not use template Haskell, and so works at both compile and run time.
We will use these ASTs to solve two problems with floating point numbers:
\begin{enumerate}
\item
We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
%This lets one programmer write naive floating point code;
%then an expert on numerical analysis can write code that automatically improves the naive code.
\item
We will specify type class laws for floating point arithmetic that can be automatically tested.
Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
\end{enumerate}
We demonstrate these techniques using GHC's built-in numeric class hierarchy,
but these techniques generalize to user defined class hierarchies as well.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}

@@

%\keywords
%floating point, homoiconic, type classes

@@

Imagine you're writing a machine learning library in Haskell.
%You decide to start by implementing logistic regression.
%The first step is to implement the log-logistic function
%Logistic regression is one of the most important machine learning algorithms,
%and to implement it, you'll need to implement the log-logistic function
To implement the classic logistic regression algorithm,
you begin by implementing the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
%This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic1| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so our function must be rewritten.

With some careful algebra, we can rewrite our log-logistic function in the following numerically stable form.
\begin{lstlisting}
logLogistic2 :: (Floating x, Ord x) => x -> x
logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
    where
        m = min 0 x
\end{lstlisting}
It is hardly obvious that these two functions are meant to calculate the same expression.
The second implementation not only contains many more terms,
it has a different type signature due to the use of the |min| function.

Constructing these numerically stable functions is a notoriously difficult black art.\footnote{
For some classic introductions, see \cite{goldberg1991,higham2002,hamming2012}.
}
Wouldn't it be nice to have a tool to automate this procedure?
We'll construct a simple one in Section 3.

The logistic regression algorithm operates on vector spaces;
so you're next task is to get a proper vector space implementation.
Since you plan to use the foreign function interface heavily,
you decide to represent vectors as a pointer to a memory location.
\begin{lstlisting}
newtype Vector (dim::Nat) a = Vector (ForeignPtr a)
\end{lstlisting}
The |dim| type parameter encodes the dimension of the vector space.
It makes sense to store this information at the type level because vector space operations don't change a vector's dimension.

To make our |Vector| type useful,
we must define how to perform the standard mathematical operations on it.
You begin by making it an instance of |Num|.
Since you're a careful programmer, you decide to write test cases to ensure your |Num| instance is correct.
You begin by writing a QuickCheck property to ensure that your |(+)| function is properly associative.
It works for |Int|s, but misteriously fails on |Float|s and |Double|s.
%Unfortunately, the standard Haskell toolkit is not sufficient for the challenge.

\newpage
%Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
%First, check our input to see if it would overflow the |exp| function.
%If no overflow would occur, then calculate our function as normal;
%but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
%Performing this transformation gives us:
%\begin{lstlisting}
%logLogistic2 :: (Floating x, Ord x) => x->x
%logLogistic2 x = if x < (-709)
    %then x
    %else log(1/1+exp(-x))
%\end{lstlisting}
%This new function is numerically stable, but we've introduced two new problems:

We've made our function numerically stable, but we've introduced two new problems:

@@

%\begin{problem}[specialization]
%The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
%This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
%To properly handle both |Float|s and |Double|s,
%we must write two separate functions.
%In practice, we may want to support many other numeric types as well.
%Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
%We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
%\end{problem}

These problems are not unique to the log-logistic function.
All functions over the real or complex numbers need to be transformed for floating point stability,
and all such manual transformations suffer from these problems.
This pearl presents a method for alleviating these problems.
Specifically, we will see how to write a function
%\begin{lstlisting}
%stabilize :: (Floating x, Ord x)
    %=> (x -> x) -> (x -> x)
%\end{lstlisting}
\begin{lstlisting}
stabilize :: (Floating x,Ord x) => (x->x) -> (x->x)
\end{lstlisting}
that automatically takes an input function like |logLogistic1| and outputs a numerically stable version of that function.
Unfortunately, this introduces a third problem:

@@

\section{Homogeneous Functions are homoiconic}
Consider the simple numeric class hierarchy shown in Figure \ref{code:group}.
With this class hierarchy, we can write a function that measures the degree to which the |(+)| operation obeys the associative rule.
\begin{lstlisting}
associator :: Group a => a -> a -> a -> a
associator a1 a2 a3
    = ((a1+a2)+a3) - (a1+(a2+a3))
\end{lstlisting}
We call the |associator| function \emph{homogeneous} because:
\begin{enumerate}
\item
The function's return type and all of its parameters are the same type variable.
\item
The constraints contain exactly one type class, and
that type class has exactly one parameter.
\end{enumerate}
Notice that each of the class methods for |Semigroup|, |Monoid|, and |Group| are also homogeneous.
Therefore, we will also say these classes are homogeneous.

@@

\subsection{A type class for FAlgebras}
Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
See for example \cite{swierstra2008}.
The heart of our construction is the |FAlgebra| type class, defined as
\begin{lstlisting}
class
    ( Functor (Sig alg)
    , Foldable (Sig alg)
    ) => FAlgebra (alg::Type->Constraint)
        where
    data Sig alg a
    runSig :: alg a => Sig alg a -> a
\end{lstlisting}
The |FAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
This means that instances of |FAlgebra| will be other type classes (instead of types or type constructors).
The || in |FAlgebra| stands for homogeneous.
Most of the names used in this section will be prefixed with || to differentiate them from related names used in the later heterogeneous constructions.

The easiest way to explain the |FAlgebra| class is by example.
We'll start with the |Semigroup| type class defined above.
\begin{lstlisting}
instance FAlgebra Semigroup where
    data Sig Semigroup a = Sig_plus a a
    runSig (Sig_plus a1 a2) = a1+a2
\end{lstlisting}
The |Sig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
That is, for every class method, |Sig| should have a constructor with the same number of parameters.
Since |Semigroup| has a single function that takes two parameters,
the |Sig Semigroup| data instance has a single constructor that takes two parameters.
The |runSig| function evaluates the signature.
For each constructor in |Sig|, |runSig| should apply the corresponding function to the constructor's parameters.

To complete the |FAlgebra Semigroup| instance, we must also define the following superclass instances.
\begin{lstlisting}
instance Functor (Sig Semigroup) where
    fmap f (Sig_plus a1 a2)
        = Sig_plus (f a1) (f a2)

instance Foldable (Sig Semigroup) where
    foldr f b (Sig_plus a1 a2)
        = f a2 (f a1 b)
\end{lstlisting}
As usual, there is only a single valid |Functor| instance.
There are multiple valid |Foldable| instances depending on the order in which |foldr| considers the constructors' parameters,
but for our purposes the order is not important.

Instances get slightly more complicated when the parameter class has superclasses.
In this case, the |Sig| data instance must also have a constructor for each superclass that stores the superclass's signature.
This can be seen in the |Monoid| and |Group| instances below.
%The |FAlgebra| instances of |Monoid| and |Group| demonstrate this.
\begin{lstlisting}
instance FAlgebra Monoid where
    data Sig Monoid a
        = Sig_zero
        | Sig_Semigroup_Monoid
            (Sig Semigroup a)
    runSig Sig_zero = zero
    runSig (Sig_Semigroup_Monoid s)
        = runSig s

instance FAlgebra Group where
    data Sig Group a
        = Sig_negate a
        | Sig_minus a a
        | Sig_Monoid
            (Sig_Monoid_Group a)
    runSig (Sig_negate a) = a
    runSig (Sig_minus a1 a2) = a1-a2
    runSig (Sig_Monoid_Group s)
        = runSig s
\end{lstlisting}
Notice that the data instance |Sig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
The corresponding |Functor| and |Foldable| instances have been omitted for brevity,
but can be found in the paper's git repo.

%The procedure for generating |FAlgebra| instances is purely mechanical.
%The ?? library automates the process with the |mkHomFAlgebra| template Haskell function.

@@

We can convert the signature of an FAlgebra into an AST as follows:
\begin{lstlisting}
type AST alg a = Free (Sig alg) a

data Free f a
    = Pure a
    | Free (f (Free f a))
\end{lstlisting}
Intuitively,
the |Pure| constructor represents a leaf in the AST,
and the |Free| constructor represents a branch.
The parameter to |Free| is filled by the appropriate |Sig| data instance.
The constructor used for the |Sig| instance corresponds to an operation,
and the parameters to the constructor will contain ASTs nested recursively.

For example, consider the expression |(1+2)+3|.
We can create a corresponding AST using |Semigroup|'s |FAlgebra| instance.
\begin{lstlisting}
Free
  (Sig_plus
    (Free
      (Sig_plus
        (Pure 1)
        (Pure 2)
      )
    )
    (Pure 3)
  )
\end{lstlisting}
We can also create a corresponding AST for any class that has |Semigroup| as a superclass.
Below is the |Group| AST.
\begin{lstlisting}
Free
  (Sig_Monoid_Group
    (Sig_Semigroup_Monoid
      (Sig_plus
        (Free
          (Sig_Monoid_Group
            (Sig_Semigroup_Monoid
              (Sig_plus
                (Pure 1)
                (Pure 2)
              )
            )
          )
        )
        (Pure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |Sig_plus| constructor must be embedded into the |Group|'s |Sig| with calls to |Sig_Monoid_Group| and |Sig_Semigroup_Monoid|.

Constructing these ASTs by hand is a tedious process.
There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |Sig_plus| constructor.
We can avoid this boilerplate with the following type class.
\begin{lstlisting}
class
    ( FAlgebra alg1
    , FAlgebra alg2
    ) => View alg1 alg2
        where
    embedHomSig
        :: Sig alg1 a -> Sig alg2 a
    unsafeExtractHomSig
        :: Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Instances of the |View| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
For example, |Semigroup| is a superclass of |Group|,
so we should have a |View Semigroup Group| instance.
\begin{lstlisting}
instance View Semigroup Group
    embedHomSig s
        = Sig_Monoid_Group (embedHomSig s)
    unsafeExtractHomSig (Sig_Monoid_Group s)
        = unsafeExtractHomSig s
\end{lstlisting}
Because of this property, we know that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance.
But it is only sometimes true that we can extract a |Sig alg1 a|,
which is why the corresponding function is labeled |unsafe|.

Armed with the |View| class, we can create a uniform representation for all class's ASTs.
The expression |(1+2)+3| can now be represented as:
\begin{lstlisting}
Free
  (embedHomSig
    (Sig_plus
      (Free
        (embedHomSig
          (Sig_plus
            (Pure 1)
            (Pure 2)
          )
        )
      )
      (Pure 3)
    )
  )
\end{lstlisting}

%Every FAlgebra can be embedded into itself.
%\begin{lstlisting}
%instance View alg alg where
    %embedSig = id
    %unsafeExtractSig = id
%\end{lstlisting}

@@

Using our uniform representation of an AST,
we can make the AST itself an instance of any homogeneous type class.
Again, we'll approach the procedure by example.
We can write a |Semigroup| instance for |AST Semigroup a| as:
\begin{lstlisting}
instance
    ( View Semigroup alg
    ) => Semigroup (Free (Sig alg) a)
        where
    (+) e1 e2 = Free $ embedHomSig $
            Sig_plus e1 e2
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedHomSig $|.
The |Monoid| and |Group| instances follow a similar pattern.
\begin{lstlisting}
instance
    ( View Semigroup alg
    , View Monoid alg
    ) => Monoid (Free (Sig alg) a)
        where
    zero = Free $ embedHomSig $
        Sig_zero

instance
    ( View Semigroup alg
    , View Monoid alg
    , View Group alg
    ) => Group (Free (Sig alg) a)
        where
    negate e  = Free $ embedHomSig $
        Sig_negate e
    (-) e1 e2 = Free $ embedHomSig $
        Sig_minus e1 e2
\end{lstlisting}
We can easily create ASTs using these instances.
The |Semigroup| AST can be created with the expression
\begin{lstlisting}
(Pure 1+Pure 2)+Pure 3
    :: AST Semigroup a
\end{lstlisting}
And the |Group| AST can be created by just changing the type signature.
\begin{lstlisting}
(Pure 1+Pure 2)+Pure 3
    :: AST Group a
\end{lstlisting}

\subsection{Showing the AST}
Now that we can create our ASTs, we want to actually do something with them!
The simplest task is to convert them into a |String| for display.
This is an easy procedure of writing |Show| instances for the |Sig| and |Free| data types.

The are no tricks in the |Show| instance for |Semigroup|'s |Sig|.
\begin{lstlisting}
instance Show a =>
    Show (Sig Semigroup a)
        where
    show (Sig_plus a1 a2)
        = show a1++"+"++show a2
\end{lstlisting}
For each operation constructor, we simply show the parameters and the operation connecting them.
For type classes with superclasses, we also need to recursively show the superclasses as can be seen in the |Monoid| and |Group| instances below.
\begin{lstlisting}
instance Show a =>
    Show (Sig Monoid a)
        where
    show (Sig_Semigroup_Monoid s)
        = show s
    show (Sig_zero) = "zero"

instance Show a =>
    Show (Sig Group a)
        where
    show (Sig_Monoid_Group s)
        = show s
    show (Sig_negate a)
        = "negate "++show a
    show (Sig_minus a1 a2)
        = show a1++"-"++show a2
\end{lstlisting}
The |Free| type's |Show| instance is similarly straightforward..
\begin{lstlisting}
instance (Show a, Show (f (Free f a)))
    => Show (Free f a)
        where
    show (Pure a) = show a
    show (Free f) = "("++show f++")"
\end{lstlisting}
If the syntax tree contains only a single leaf, we just show that leaf.
If the syntax tree contains a branch,
then put parenthesis around the branch so that we don't lose information about the order of operations, and show the functor in the middle.

\subsection{Evaluating the syntax tree}
Another useful task for ASTs is to evaluate them,
generating the value they represent.
This data type is called the free monad because as long as the parameter |f| is a |Functor|,
then |Free| has a valid |Monad| instance.
Since all |Monad|s are by necessity also |Functor|s,
|Free| has a valid |Functor| instance as well.
\begin{lstlisting}
instance Functor f => Functor (Free f) where
    fmap g (Pure a) = Pure (g a)
    fmap g (Free f)
        = Free (fmap (fmap g) f)
\end{lstlisting}
For our purposes, that's all the structure we'll need.

First, we construct a function that ``runs'' the syntax tree, converting it into a single value.
\begin{lstlisting}
runAST :: (FAlgebra alg, alg a)
    => AST alg a -> a
runAST (Pure a) = a
runAST (Free f) = runSig (fmap evalHom f)
\end{lstlisting}
If our syntax tree already consists of just a single value (i.e. it is the |ePure| constructor),
then we just return that value.
Otherwise, we use the |Functor| instance to recursively convert the |Sig|s into types of |a|, then a call to |runSig| combines the results.

\subsection{Homoiconicity at last!}
We are finally ready to see that our construction makes homogeneous functions homoiconic.
If we open GHCi and enter the following expression
\begin{lstlisting}
ghci> let ast=associator (Pure a1) (Pure a2) (Pure a3)
\end{lstlisting}
the display shows our AST
\begin{lstlisting}
\end{lstlisting}
We can also evaluate our AST and see that doing so gives the same results as running the original expression.
\begin{lstlisting}
ghci> runAST ast
